---
title: "Final"
output: pdf_document
---


```{r}
library(tsdl)
ozo.ts = tsdl[[334]]
plot.ts(ozo.ts)
n = length(ozo.ts)
fits = lm(ozo.ts ~ as.numeric(1:n));abline(h=mean(ozo.ts), col = 'blue');abline(fits, col="red")
```


Next
```{r}
ozo = ozo.ts[c(1:204)] #truncating data
ozo.test = ozo.ts[c(205:216)]
plot.ts(ozo) #plotting new truncated data
lm(ozo ~ as.numeric(1:length(ozo)))
fit = lm(ozo ~ as.numeric(1:length(ozo))); abline(fit, col="red")
abline(h=mean(ozo), col="blue")
```



Next
```{r}
par(mfrow = c(2,2))
hist(ozo, col="light blue", xlab="", main="histogram; ozo data") #Skewed Data
acf(ozo,lag.max=40, main="ACF of the ozo data") #ACF large and periodic
```



Next

```{r}
library(MASS)
bcTransform = boxcox(ozo~ as.numeric(1:length(ozo)))  #plots the graph

lambda = bcTransform$x[which(bcTransform$y == max(bcTransform$y))] #gives lambda

ozo.bc = (1/lambda)*(ozo^lambda-1)  #bc transform
ozo.log = log(ozo)    #log transform
ozo.s = sqrt(ozo)   #sqrt transform
plot.ts(ozo.bc, main = 'ozo boxcox transform')
plot.ts(ozo.log, main = 'ozo log transform')

plot.ts(ozo.s, , main = 'ozo square_root transform')

hist(ozo.log, col="light blue", xlab="", main="histogram; ozo log transform")
hist(ozo.bc, col="light blue", xlab="", main="histogram; ozo bc transform") 

hist(ozo.s, col="light blue", xlab="", main="histogram; ozo sqrt transform") 
var(ozo.s)   #variance of sqrt, log and bc
var(ozo.log)
var(ozo.bc)
```


next

```{r}
library(ggplot2) 
library(ggfortify) 
install.packages("ggplot2")
install.packages("ggfortify")
y = ts(as.ts(ozo.s), frequency = 12)
decomp = decompose(y)  #Decomposition of the series
plot(decomp) #plotting the decomp
```


Next
```{r}
var(ozo.s) #var is 0.001144

ozo.s_12 = diff(ozo.s, lag=12) #differenced at lag 12

plot.ts(ozo.s_12, main="s differenced at lag 12")
var(ozo.s_12)  #variance is close to 0
fit = lm(ozo.s_12 ~ as.numeric(1:length(ozo.s_12))); abline(fit, col="red")
mean(ozo.s_12)  #mean is close to 0
abline(h=mean(ozo.s_12), col="blue")

ozo.dif = diff(ozo.s_12, lag=1) #differenced at lag 12 and 1
plot.ts(ozo.dif, main="s differenced at lag 12 & lag 1")
fit = lm(ozo.dif ~ as.numeric(1:length(ozo.dif))); abline(fit, col="red") 
mean(ozo.dif) #mean is closer to 0
abline(h=mean(ozo.dif), col="blue")
var(ozo.dif)  #variance is closer to 0
```


NExt
```{r}
acf(ozo.s_12, lag.max=40, main="ACF of the squareroot, differenced at lag 12")
acf(ozo.dif, lag.max=40, main="ACF of the squareroot, differenced at lags 12 and 1")
hist(ozo.dif, col="light blue", xlab="", main="histogram; squareroot differenced at lags 12 & 1")
```

Next

```{r}
acf(ozo.dif,40,main="ACF of the squareroot_transform, differenced at lags 12 and 1") #shows signifance at lags 1,2
pacf(ozo.dif,40,main="PACF of the squareroot_transform, differenced at lags 12 and 1") #shows decay
```


Possible models MA q = 1 or 2. Introduce AR components to see what happens to AIC


```{r}

arima(ozo.s, order=c(0,1,1), seasonal = list(order = c(0,1,1), period = 12), method="ML") # p = 0, q = 1

arima(ozo.s, order=c(0,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML") # p = 0, q = 2

arima(ozo.s, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML") # p = 0, q = 3



arima(ozo.s, order=c(1,1,1), seasonal = list(order = c(0,1,1), period = 12), method="ML") # p = 1, q = 1

arima(ozo.s, order=c(1,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML") # p = 1, q = 2

#fix as third coefficient is 0
arima(ozo.s, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), fixed = c(NA, NA, 0, NA), method="ML")

#fix as 1 and 3 coefs are 0
arima(ozo.s, order=c(1,1,2), seasonal = list(order = c(0,1,1), period = 12), fixed = c(0, NA, 0, NA), method="ML")

```
Confirm with AICc function

```{r}
AICc(arima(ozo.s, order=c(0,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")) #-41.6
AICc(arima(ozo.s, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), fixed = c(NA, NA, 0, NA), #-41.6 method="ML"))
AICc(arima(ozo.s, order=c(1,1,1), seasonal = list(order = c(0,1,1), period = 12), method="ML"))
```

Choose top 3 models with lowest AIC

Diagnostics for Model 2 which is arima(0,1,2)(0,1,1)12
```{r}
fit = arima(ozo.s, order=c(0,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
res_1 = residuals(fit) #residuals from model 2
hist(res_1,density=20,breaks=20, col="blue", xlab="", prob=TRUE) 
m = mean(res_1) #mean close to 0
std = sqrt(var(res_1))
curve( dnorm(x,m,std), add=TRUE )
plot.ts(res_1)
fitt = lm(res_1 ~ as.numeric(1:length(res_1))); abline(fitt, col="red") 
abline(h=mean(res_1), col="blue")
qqnorm(res_1,main= "Normal Q-Q Plot for Model B")
qqline(res_1,col="blue")
acf(res_1, lag.max=40)
pacf(res_1, lag.max=40)
#Stats Tests
shapiro.test(res_1)
Box.test(res_1, lag = 12, type = c("Box-Pierce"), fitdf = 2)
Box.test(res_1, lag = 12, type = c("Ljung-Box"), fitdf = 2)
Box.test(res_1^2, lag = 12, type = c("Ljung-Box"), fitdf = 0)
acf(res_1^2, lag.max=40)
ar(res_1, aic = TRUE, order.max = NULL, method = c("yule-walker"))
```


Diagnostics for model 4 which is ARIMA(1,1,1)(0,1,1)
par(mfrow = c(2,2))
```{r}
fit_2 = arima(ozo.s, order=c(1,1,1), seasonal = list(order = c(0,1,1), period = 12), method="ML")
res_2 = residuals(fit_2) #residuals from model 4

hist(res_2,density=20,breaks=20, col="blue", xlab="", prob=TRUE)
m_2 = mean(res_2) #mean close to 0
std_2 = sqrt(var(res_2))
curve( dnorm(x,m_2,std_2), add=TRUE )
plot.ts(res_2)
fitt_2 = lm(res_2 ~ as.numeric(1:length(res_2))); abline(fitt_2, col="red") 
abline(h=mean(res_2), col="blue")
qqnorm(res_2,main= "Normal Q-Q Plot for Model B")
qqline(res_2,col="blue")
acf(res_2, lag.max=40)
pacf(res_2, lag.max=40)
#Stats Tests
shapiro.test(res_2)
Box.test(res_2, lag = 12, type = c("Box-Pierce"), fitdf = 3)
Box.test(res_2, lag = 12, type = c("Ljung-Box"), fitdf = 3)
Box.test(res_2^2, lag = 12, type = c("Ljung-Box"), fitdf = 0)
acf(res_2^2, lag.max=40)
ar(res_2, aic = TRUE, order.max = NULL, method = c("yule-walker"))
```

Diagnostics for model 3, which is Arima(0,1,3)(0,1,1) where coeffs are fixed

```{r}
fit_3 = arima(ozo.s, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), fixed = c(NA, NA, 0, NA), method="ML")
res_3 = residuals(fit_3) #residuals from model 3

hist(res_3,density=20,breaks=20, col="blue", xlab="", prob=TRUE)
m_3 = mean(res_3) #mean close to 0
std_3 = sqrt(var(res_3))
curve( dnorm(x,m_3,std_3), add=TRUE )
plot.ts(res_3)
fitt_3 = lm(res_3 ~ as.numeric(1:length(res_3))); abline(fitt_3, col="red") 
abline(h=mean(res_3), col="blue")
qqnorm(res_3,main= "Normal Q-Q Plot for Model B")
qqline(res_3,col="blue")
acf(res_3, lag.max=40)
pacf(res_3, lag.max=40)
#Stats Tests
shapiro.test(res_3)
Box.test(res_3, lag = 12, type = c("Box-Pierce"), fitdf = 3)
Box.test(res_3, lag = 12, type = c("Ljung-Box"), fitdf = 3)
Box.test(res_3^2, lag = 12, type = c("Ljung-Box"), fitdf = 0)
acf(res_3^2, lag.max=40)
ar(res_3, aic = TRUE, order.max = NULL, method = c("yule-walker"))
```



Model 1 is difionary cuz it is pure MA
Check Invertibility

```{r}
par(mfrow = c(2,2))
plot.roots(NULL,polyroot(c(1, -0.6649, -0.1991)), main="Model 2, roots for MA part") #Roots for model 2
plot.roots(NULL,polyroot(c(1, -0.9010, -0.252, 0)), main="Model 3, roots for MA part ") #rootds for model 3
plot.roots(NULL,polyroot(c(1, 0.2265)), main="Model 4, roots for AR part ") #roots for model 4
plot.roots(NULL,polyroot(c(1,-0.9026)), main="Model 4, roots for MA part ") #roots for model 4

```

So Model 1 is also invertible as the roots are outside unit circle

For Model 2, difionarity and invertibility must be checked as it is ARMA.

```{r}
polyroot(c(1, 0.2265))
polyroot(c(1,-0.9026))
          
```

Model 2 is invertible and difionary


Model 3 is difionary cause it is pure MA
Check Invertibility
```{r}
polyroot(c(1, -0.9010, -0.252, 0))
```



```{r}
library(forecast)
fit.A = arima(ozo.s, order=c(0,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
forecast(fit.A) #forecasting model 2

pred.tr = predict(fit.A, n.ahead = 12) #creating prediction
U.tr= pred.tr$pred + 2*pred.tr$se #95% CI upper bounds
L.tr= pred.tr$pred - 2*pred.tr$se #95% CI lower bounds
ts.plot(ozo.s, xlim=c(1,length(ozo.s)+12), ylim = c(min(ozo.s),max(U.tr)))
lines(U.tr, col="blue", lty="dashed")
lines(L.tr, col="blue", lty="dashed")
points((length(ozo.s)+1):(length(ozo.s)+12), pred.tr$pred, col="red")


```


```{r, figures-side, fig.show="hold", out.width="50%"}
pred.orig = pred.tr$pred**2 #squaring data to recreate original data
U= (U.tr)**2 #squaring CI's
L= (L.tr)**2
ts.plot(ozo, xlim=c(1,length(ozo)+12), ylim = c(min(ozo),max(U)), ylab = 'Forecast Ozone')
lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")
points((length(ozo)+1):(length(ozo)+12), pred.orig, col="red")
lines(ozo.ts, col = 'red')

```

```{r}
ts.plot(ozo.ts, xlim=c(1955,1973), ylim = c(1,5), ylab = 'Original Data')
```
